Repeatability of Pre-Flashover Fire Patterns on Gypsum Wallboard

### 2016-10-04 Projects/Madrzykowski_Thesis/Report/Madrzykowski_Thesis.tex
FDS text

List of Symbols
\begin{tabbing}
\hspace{1.5in} \= \\
$c_{p}$ \> specific heat \\
$k$ \> thermal conductivity \\
$\chi_r$ \> radiative fraction \\
$\dot{q}''$ \> heat flux per unit area kW/m$^2$ \\
$\dot{Q}$ \> heat release rate (kW) \\ 
$R$ \> distance from the point source to the target (m) \\
$\rho$ \> density \\
\end{tabbing}

@@

This section reviews the fire problem to demonstrate the need and value to society that fire investigation can provide. Following the review, a brief presentation on the analysis  of fire patterns, the use of analytical tools, and the current challenges to the field of fire forensics is given as they are the reason for this study.  In closing this section, an example of a fire incident is presented. In the closing section we will return to the example again.

\subsection{The Need for Fire Investigation}
Unwanted fires result in loss of life, property and can create an adverse economic impact on a community.  Recent findings from an archaeological site in South Africa, indicate that mankind has productively used fire for approximately 1 million years~\cite{Berna:2012}.  This is 300 thousand years earlier than previously thought.  Despite mankind's extensive history with fire, according to the World Health Organization, an estimated 265,000 people die each year as a result of burn injuries~\cite{WHO:2014}.  Based on an estimated global population of 7.3 billion people, the death rate due to burns from all sources would be 3.6 per 100,000 people.  The World Fire Statistics Center collects data from European, North American and a few Asia-Oceania countries.  Within this group of countries the fire deaths per 100,000 population range from 0.02 (Singapore) to 2.03 (Finland)~\cite{Climate:2014}. There are a significant number of factors that determine the variability of fire death rates between nations, socio-economic conditions being a leading factor~\cite{WHO:2014}.

In the United States, the loss of life due to unwanted fire has been decreasing steadily since the 1970’s.  This trend does not include loss of life as a result of the terrorist attack on September 11, 2001.  The estimated number of fire deaths in the U.S. in 1975 was 8,100 or 3.75 deaths per 100,000 population~\cite{America_Burning_Revisited}.  In 2013, there were 3,240 civilian fire fatalities or 1.02 deaths per 100,000 population~\cite{Karter:2014}.  During this same period, the number of fires in the United States has decreased from 2.5 million fires to 1.24 million fires.  There are many reasons that led to this decrease in loss of life due to fire in the United States.  An example would be research and data that enabled the development of product safety standards, improved building codes, improved fire safety systems, and fire prevention programs to name just a few.  None of this happened without the constant effort of agencies at the Federal, state, and, local levels working with professional fire organizations to deliver the five E’s of fire prevention; Engineering, Education, Enforcement, Economic Incentive, and Emergency Response~\cite{FEMA:2013}.

Fire investigation is a key element in the fire prevention system that generates data to enable the reduction of fire losses.  The investigation of fires provides a means to identify the cause of the fire in order to develop a knowledge base that could enable the elimination of that cause and thus further reduce the losses from fires.  Data such as the room of fire origin, the first item ignited and ultimately the cause of the fire is information that is critical to understanding and reducing the number of fires.  For example, the identification of products that are involved in ignitions due to a design flaw, in many cases result in the U.S. Consumer Product Safety Commission or a manufacturer issuing a product recall.

In some cases, the fire investigation determines that the fire was intentional.  An intentional fire as defined in the National Fire Incident Reporting System (NFIRS)  are those fires that are deliberately set and include fires that result from deliberate misuse of a heat source, fires of an incendiary nature (arson), as well as controlled burn fires, such as crop clearing, that required fire service intervention~\cite{Campbell:2014}.  An incendiary fire, as defined by the National Fire Protection Association (NFPA) Guide for Fire \& Explosion Investigations, commonly referred to as NFPA~921, is ``a fire that is deliberately set with the intent to cause a fire to occur in an area where the fire should not be''~\cite{NFPA:921}.

@@

The values cited above are believed to be the best available in the United States and perhaps one of the better fire incident databases in the world.  However an investigative report by Icove and Hargrove, indicates that arson fires are underreported to the U.S. National Fire Incident Reporting System (NFIRS).  The investigation provided examples from several major cities, such as New York City, where in 2011, 11 arson fires in buildings were reported to NFIRS, while the Fire Department of the City of New York's Bureau of Fire Investigation had recorded 1,347 arson fires in buildings that year.  Similar discrepancies were found in eight other large U.S. cities.  On average, three-fourths of the arsons uncovered by city fire investigators went unreported to NFIRS~\cite{Icove_2014}.  In other words, the number of arson related incidents is likely to be under represented in the fire incident data.

\subsection{The Use of Fire Patterns in Fire Investigation}
Fires are investigated in order to determine the ``origin and cause'' of the fire and in cases of arson there is an effort to determine who was responsible for setting the fire.  This investigation process must follow the scientific method as documented in NFPA 921~\cite{NFPA:921}.  There are numerous textbooks on the subject of conducting a fire investigation~\cite{Almirall:2004,Fire_Investigation,DeHaan:2012,Icove:2013,Lentini:2006,Noon:1995}. The NFPA guide and the texts provide information how to collect data from the scene, analyze the data, and develop a hypothesis about the fire.

@@

Previous fire pattern research by the National Institute of Justice (NIJ), the National Institute of Standards and Technology (NIST), and the United States Fire Administration (USFA) has shown that fire patterns provide data useful for the determination of fire origin.  The reports noted the impact of ventilation on the development of the burn patterns~\cite{Shanley:1997,Putorti:1997}. A large number of other factors affect the formation of these patterns: burn time, heat release rate of the fire source, fire exposure, target fuel composition, adjacent fuel(s) and compartmentalization, to name a few. Given the limited number of experiments in the literature and the large number of variables, it has been difficult to develop a cause and effect relationship between the fire scenarios and the resulting patterns, based solely on the research.

\subsection{The Use of Fire Models in Fire Investigation}
Fire models are another tool that may be used by a fire investigation team to gain insight into the growth and development of a fire and to test the investigator’s hypothesis on origin and cause~\cite{Sutula:2001}.  Many theoretical and empirical correlations and computational fire models have been developed for use in fire protection design, fire hazard analysis, and fire investigation.  A collection of the widely used correlations can be found in the U.S. Nuclear Regulatory Commission’s (NRC) Fire Dynamics Tools (FDTs). These spread sheets incorporate references to the source documents and experimental work upon which the correlations are based. In addition, NRC tasked NIST with conducting a verification and validation study of the correlations. Overholt developed the NIST verification and validation study~\cite{Overholt:2014}. Not all of the correlations in the FDTs were included in the study because an insufficient amount of reliable data is available to conduct validation.  The correlations for flame height and time to ignition were not included in the validation study.

In 2003, a list of zone and computational fluid dynamics (CFD) fire models was compiled by Olenick and Carpenter~\cite{Olenick:2003}. A similar survey was conducted in 1992, and Olenick updated the list in 2014. There are 53 zone models listed, but only three are actively supported or maintained.  The survey also includes 22 field models of which 10 are listed as being actively supported.  All of the correlations and models have, to some extent, been validated or checked against experimental data for a limited set of scenarios~\cite{ASTM_E1355}.

@@

However for fire models to be demonstrated as reliable and suitable for use in court, studies such as the one conducted here need to be conducted to provide the scientific basis for the use of fire models for re-construction.  The research results may enable fire investigation teams to have a better understanding of the fire scene and enable improved support of their hypothesis on the area of origin based on fire patterns.  This idea was discussed when the Fire Protection Research Foundation convened a Research Advisory Council on Post-fire Analysis in 2002.  Recommendations for research and development in their white paper included, ``advance the capabilities of computational fluid dynamics fire modeling, particularly as applied to fire ignition scenarios and fire pattern development and interpretation''~\cite{RAC:2002}.

Given that fire pattern interpretation and analysis is a critical step in most fire investigations, it is surprising to find the limited amount of research and analysis tools directly related to this practice.  Gorbett recently conducted a review of fire pattern research that has been conducted over the last 80 years. The review points out many gaps, especially with how fire patterns are used to determine the area of origin.  A key limitation is the assumption by many of the researchers, or text authors, that investigators are able to visibly assess varying degrees of fire damage and then determine the area of origin~\cite{Gorbett_2015}. Studies have shown that investigators have not been able to use only the location of the fire patterns as means of reliably determining the area of origin, some amount of analysis and fire dynamics knowledge is required~\cite{Carmen_2008},~\cite{Tinsley_2013}.

@@

As the result of the review of several arson cases, most notably the cases of Ernest Ray Willis and Cameron Todd Willingham, the practice of fire investigation as a forensic science was called into question.  Beyler conducted a review of both cases for the Texas Forensic Science Commission.  Both fires occurred prior to the release of NFPA~921 in 1992.  The Beyler analysis, issued in August of 2009, concluded that the investigations of both cases ``did not comport with either the modern standard of care expressed by NFPA 921, or the standard of care expressed by fire investigation texts and papers in the period 1980 {\textendash} 1992.'' The fire investigators had a poor understanding of science and their findings of arson could not be sustained~\cite{Beyler:2009}.

@@

Given that the majority of intentional fires in structures occurred in residential occupancies, this study will focus on fire patterns on painted gypsum wallboard which is a common interior finish for residential occupancies.  A common height for a residential occupancies is 2.4~m in height.  With a wall height of 2.4~m, these experiments will use small fires with mean flame heights of approximately 1.2~m.  The resulting patterns on the gypsum board wall would be representative of a pre-flashover fire pattern.

@@

This study includes several series of real-scale, replicate fire experiments to examine the repeatability of pre-flashover fire patterns, and ultimately examine the ability of simple correlations and the Fire Dynamic Simulator to recreate characteristics of the fire patterns.

@@

Statistical methods were used to guide the wall fire and corner fire experiments.  The analysis ensured that the appropriate number of replicate experiments were conducted to provide an assessment of burn pattern repeatability within a given level of uncertainty.  Given that the coefficient of variation (COV) may be different for a given type of experiment, the number of replicates required would be larger for experiments with a larger COV in order obtain a valid assessment of repeatability.

@@

A free standing, steel frame with non-combustible, calcium silicate faced wall was instrumented to measure the thermal exposure of the wall from the fire in terms of temperature and heat flux.  The fires were initially placed at predetermined, fixed positions away from the wall. The burner was moved toward the wall until the burner was against the wall.

@@

A free standing, steel frame with a panel of gypsum wallboard attached to the surface was instrumented to measure the thermal exposure of the wall from the fire in terms of temperature and heat flux.  The fires were initially placed at predetermined, fixed positions away from the wall. The burner was moved toward the wall until the burner was against the wall.  If the results indicated that no burn pattern would result from a given combination of fuel and position, then that fuel and position combination were eliminated from the remainder of the test matrix.

@@

These experiments used the same test compartment as the wall fire pattern experiments.  In these cases the fire was located against both walls in a corner of the compartment.  The experimental process was the same as the wall fire pattern experiments.

@@

Cone calorimeter, ASTM~E1354, experiments were conducted to estimate the ignition temperature, the time to ignition and the heat release rate from samples of painted gypsum wallboard under thermal flux conditions ranging from 25~kW/m$^2$ to 75~kW/m$^2$ ~\cite{ASTM_E1354}.

@@

FDS~\cite{McGrattan:2014} and Smokeview~\cite{Forney:2014} were used to examine the ability to numerically reproduce the source fires and the instrumented wall experiments. These simulation tools provide results in the form of visual field data, in other words they provide results in the same format that the investigator would see or document the fire pattern.  Comparison between the experiments and the model may provide valuable validation data for the model.

@@

Point measurements were reduced and compared with values from the literature on fire plumes and their interaction with vertical surfaces.  The heat release rates, heat flux measurements, flame and surface temperatures were used as input for heat transfer analysis to develop empirical coefficients for the specific experimental arrangements which would be useful in the model recreations.
The key measurements for repeatability of the fire patterns on the gypsum wallboard are height, maximum width, shape, and area.  The two significant outcomes will be the repeatability of the fire patterns themselves and the ability to numerically recreate key measures of the pattern.

@@

A series of experiments were conducted to characterize a limited set of  residential scale, pre-flashover or early growth stage fires.  A fixed fire source footprint, 30.5~cm square, was chosen for this study and a range of common fuels that could be found in and around a residence were selected.  The fires were positioned under an oxygen consumption calorimetry hood to determine the heat release rate.  Heat release rate, flame/plume gas temperatures, and total heat flux from the flames were measured and flame movement and height were recorded with photographs and videos.  Given the focus on residential fires, small source fires were selected so that the burn pattern would be limited to less than 2.4~m in height.  Replicate experiments were conducted with each of the fuels, in order to examine the repeatability of the fires and to quantify the variability in the flames.

@@

Three different fuels were used in this study.  Experiments were conducted with a natural gas fueled burner, a liquid fueled pan fire, and a solid fuel. Natural gas was chosen as a fuel for two reasons; it is a common fuel used for heating homes and it is used for calibrating the oxygen depletion calorimeters in NIST’s Large Fire Laboratory.  Therefore the heat content of the gas was monitored and mass flow controlled to provide a heat release rate based on fuel mass flow for comparison with the heat release rate values measured with the calorimeter.  The gas burner arrangement provides the most idealized fire, in that it can be ignited and brought up to a near steady state heat release rate within a few seconds and then be shut down just as quickly.  This provided an approximate `square wave' exposure to the gypsum wallboard.  This is important for determining the potential amount of energy transferred to the exposed surface.

The natural gas burner was 30.5~cm on a side and the top surface of the burner was 7.5~cm above the floor. The shell of the burner was made from steel and filled with ``pea sized'' gravel with an average diameter of 6~mm.  A steady state heat release rate of approximately 80 kW was used for the experiments.

Gasoline was the fuel of choice for the liquid fueled pan fire.  Gasoline is not an ideal fuel to use in experiments because it is a mixture of components and additives that are specific to manufacturers. In addition the mixtures change from season to season, and vary based on requirements or restrictions of the locality.  However, fire incident data shows that gasoline was the first item ignited in the majority of intentional fires where flammable or combustible liquids were used~\cite{Rohr:2001}. In addition, data from forensic laboratories, collected by Babrauskas and others, indicates that gasoline is the most prevalent accelerant found as part of the analysis of fire debris samples~\cite{Babrauskas:2003,Chasteen:2010}.

The gasoline pan fire used 500~mL of unleaded, 87 octane gasoline.  The pan was constructed from 6~mm thick steel.  The inner dimension of the pan was 30.5~cm on a side and 1.9~cm high.  The pan was elevated in order to bring the pan lip height up to 7.5~cm above the floor to be the same height as the natural gas burner.  The 500~mL of gasoline had a fuel depth of 5.5~mm and provided a burn time of approximately 300~s.

Flexible polyurethane foam was chosen as the solid fuel. According to the Polyurethane Foam Association, flexible polyurethane foam is the most widely used cushioning material in upholstered furnishing and mattresses. More than 1.2 billion pounds are produced in the U.S. every year~\cite{Foam,Polyurethane_Foam}. In studies by Ahrens and Evarts of the NFPA, upholstered furniture and mattresses and bedding were the first items ignited in 33~\% of the fatal fires in the U.S. based on averaged data from 2005 through 2009~\cite{Ahrens:2011,Evarts:2011}.

The flexible polyurethane foam had a density of approximately 23~kg/m$^3$.  Each block of foam was 30.5~cm on a side and was 7.5~cm thick. The bottom and sides of the foam block were wrapped in aluminum foil to prevent the fuel from flowing or spreading as it burned to ensure that the fuel area remained with the same fixed area as the gasoline and natural gas.

@@

On each fuel’s averaged heat release rate curve, the shaded area surrounding the average represents the 95~\% confidence level (2$\sigma$), based on Type A evaluation of the uncertainty is provided~\cite{Taylor:1994}.  A Type A evaluation of the uncertainty is any statistical analysis of the data from a series of observations.  In this case the evaluation assumed a normal distribution of the data such that doubling the calculated standard deviation (2$\sigma$) would yield a confidence level of approximately 95~\%.

@@

@@

>>>>

List of Symbols
\begin{tabbing}
\hspace{1.5in} \= \\
$c_{p}$ \> specific heat \\
$k$ \> thermal conductivity \\
$\chi_r$ \> radiative fraction \\
discusion$\dot{q}''$ \> heat flux per unit area kW/m$^2$ \\
$\dot{Q}$ \> heat release rate (kW) \\ 
$R$ \> distance from the point source to the target (m) \\
$\rho$ \> density \\
\end{tabbing}

@@

This section reviews the fire problem to demonstrate the need and value to society that fire investigation can provide. Following the review, a brief presentation on the analysis  of fire patterns, the use of analytical tools, and the current challenges to the field of fire forensics is given as they are the reason for this study.  At the end of this chapter, an example of a fire incident is presented. In Chapter 9, we will return to the example again.

\subsection{The Need for Fire Investigation}
Unwanted fires result in loss of life, property, and can create an adverse economic impact on a community.  Recent findings from an archaeological site in South Africa, indicate that mankind has productively used fire for approximately 1 million years~\cite{Berna:2012}.  This is 300 thousand years earlier than previously thought.  Despite mankind's extensive history with fire, according to the World Health Organization, an estimated 265,000 people die each year as a result of burn injuries~\cite{WHO:2014}.  Based on an estimated global population of 7.3 billion people, the death rate due to burns from all sources would be 3.6 per 100,000 people.  The World Fire Statistics Center collects data from European, North American and a few Asia-Oceania countries.  Within this group of countries the fire deaths per 100,000 population range from 0.02 (Singapore) to 2.03 (Finland)~\cite{Climate:2014}. There are a significant number of factors that determine the variability of fire death rates between nations, socio-economic conditions being a leading factor~\cite{WHO:2014}.

In the United States, the loss of life due to unwanted fire has been decreasing steadily since the 1970’s.  This trend does not include loss of life as a result of the terrorist attack on September 11, 2001.  The estimated number of fire deaths in the U.S. in 1975 was 8,100, or 3.75 deaths per 100,000 population~\cite{America_Burning_Revisited}.  In 2013, there were 3,240 civilian fire fatalities, or 1.02 deaths per 100,000 population~\cite{Karter:2014}.  During this same period, the number of fires in the United States has decreased from 2.5 million fires to 1.24 million fires.  There are many reasons that led to this decrease in loss of life due to fire in the United States.  One example is research and data that enabled the development of product safety standards, improved building codes, improved fire safety systems, and fire prevention programs.  None of this happened without the constant effort of agencies at the Federal, state, and, local levels working with professional fire organizations in order to deliver the five E’s of fire prevention; Engineering, Education, Enforcement, Economic Incentive, and Emergency Response~\cite{FEMA:2013}.

Fire investigation is a key element in the fire prevention system that generates data to enable the reduction of fire losses.  The investigation of fires provides a means to identify the cause of the fire in order to develop a knowledge base that could enable the elimination of that cause, and thus further reduce the losses from fires.  Data such as the room of fire origin, the first item ignited, and ultimately the cause of the fire is information that is critical to understanding and reducing the number of fires.  For example, the identification of products that are involved in ignitions due to a design flaw in many cases result in the U.S. Consumer Product Safety Commission or a manufacturer issuing a product recall.

In some cases, the fire investigation determines that the fire was intentional.  An intentional fire, as defined in the National Fire Incident Reporting System (NFIRS), are those fires that are deliberately set and include fires that result from deliberate misuse of a heat source, fires of an incendiary nature (arson), as well as controlled burn fires, such as crop clearing, that required fire service intervention~\cite{Campbell:2014}.  An incendiary fire, as defined by the National Fire Protection Association (NFPA) Guide for Fire \& Explosion Investigations, commonly referred to as NFPA~921, is ``a fire that is deliberately set with the intent to cause a fire to occur in an area where the fire should not be''~\cite{NFPA:921}.

@@

The statistics cited above are believed to be the best available in the United States, which is perhaps one of the better fire incident databases in the world.  However, an investigative report by Icove and Hargrove indicates that arson fires are underreported to the U.S. National Fire Incident Reporting System (NFIRS).  The investigation provided examples from several major cities, such as New York City. In 2011, 11 arson fires in buildings were reported to NFIRS, while the Fire Department of the City of New York's Bureau of Fire Investigation recorded 1,347 arson fires in buildings that year.  Similar discrepancies were found in eight other large U.S. cities.  On average, three-fourths of the arsons uncovered by city fire investigators went unreported to NFIRS~\cite{Icove_2014}.  In other words, the number of arson related incidents is likely to be under represented in the fire incident data.

\subsection{The Use of Fire Patterns in Fire Investigation}
Fires are investigated in order to determine the ``origin and cause'' of the fire.  In cases of arson, there is an additional effort to determine who was responsible for setting the fire.  This investigation process must follow the scientific method, as documented in NFPA 921~\cite{NFPA:921}.  There are numerous textbooks on the subject of conducting a fire investigation~\cite{Almirall:2004,Fire_Investigation,DeHaan:2012,Icove:2013,Lentini:2006,Noon:1995}. The NFPA guide and the texts provide information how to collect data from the scene, analyze the data, and develop a hypothesis about the fire.

@@

Previous fire pattern research by the National Institute of Justice (NIJ), the National Institute of Standards and Technology (NIST), and the United States Fire Administration (USFA) has shown that fire patterns provide data useful for the determination of fire origin.  The reports noted the impact of ventilation on the development of the burn patterns~\cite{Shanley:1997,Putorti:1997}. A large number of other factors affect the formation of these patterns: burn time, heat release rate of the fire source, fire exposure, target fuel composition, adjacent fuel(s), and compartmentalization, to name a few. Given the limited number of experiments in the literature and the large number of variables, it has been difficult to develop a cause and effect relationship between the fire scenarios and the resulting patterns, based solely on the research.

\subsection{The Use of Fire Models in Fire Investigation}
Fire models are another tool that may be used by a fire investigation team to gain insight into the growth and development of a fire and to test the investigator’s hypothesis on origin and cause~\cite{Sutula:2001}.  Many theoretical and empirical correlations and computational fire models have been developed for use in fire protection design, fire hazard analysis, and fire investigation.  A collection of the widely used correlations can be found in the U.S. Nuclear Regulatory Commission’s (NRC) Fire Dynamics Tools (FDTs)~\cite{NUREG_1805}. These spreadsheets incorporate references to the source documents and experimental work upon which the correlations are based. NRC tasked NIST with conducting a verification and validation study of these correlations. Overholt developed the NIST verification and validation study~\cite{Overholt:2014}. Not all of the correlations in the FDTs were included in the study, because an insufficient amount of reliable data is available to conduct validation.  The correlations for flame height and time to ignition were not included in the validation study.

In 2003, a list of zone and computational fluid dynamics (CFD) fire models was compiled by Olenick and Carpenter~\cite{Olenick:2003}. A similar survey was conducted in 1992, and Olenick updated the list in 2014. There are 53 zone models listed, but only three are actively supported or maintained.  The survey also includes 22 field models, of which 10 are listed as being actively supported.  All of the correlations and models have been validated to some extent, or checked against experimental data for a limited set of scenarios~\cite{ASTM_E1355}.

@@

In order for fire models to be demonstrated as reliable and suitable for use in court, studies such as the one conducted here need to be conducted to provide the scientific basis for the use of fire models for re-construction.  These research results may enable fire investigation teams to have a better understanding of the fire scene and enable improved support of their hypothesis on the area of origin based on fire patterns.  This idea was discussed when the Fire Protection Research Foundation convened a Research Advisory Council on Post-fire Analysis in 2002.  Recommendations for research and development in their white paper included, ``advance the capabilities of computational fluid dynamics fire modeling, particularly as applied to fire ignition scenarios and fire pattern development and interpretation''~\cite{RAC:2002}.

Since fire pattern interpretation and analysis is a critical step in most fire investigations, it is surprising to find the limited amount of research and analysis tools directly related to this practice.  Gorbett recently conducted a review of fire pattern research that has been conducted over the last 80 years. The review points out many gaps, especially with how fire patterns are used to determine the area of origin.  A key limitation is the assumption by many of the researchers, or text authors, that investigators are able to visibly assess varying degrees of fire damage and then determine the area of origin~\cite{Gorbett_2015}. Studies have shown that investigators have not been able to use only the location of the fire patterns as means of reliably determining the area of origin.  Some amount of analysis and fire dynamics knowledge is required~\cite{Carmen_2008},~\cite{Tinsley_2013}.

@@

As the result of the review of several arson cases, most notably the cases of Ernest Ray Willis and Cameron Todd Willingham, the practice of fire investigation as a forensic science was called into question.  Beyler conducted a review of both cases for the Texas Forensic Science Commission~\cite{Beyler:2009}.  Both fires occurred prior to the release of NFPA~921 in 1992.  The Beyler analysis, issued in August of 2009, concluded that the investigations of both cases ``did not comport with either the modern standard of care expressed by NFPA 921, or the standard of care expressed by fire investigation texts and papers in the period 1980 {\textendash} 1992.'' The fire investigators had a poor understanding of science and their findings of arson could not be sustained~\cite{Beyler:2009}.

@@

The majority of intentional fires in structures have occurred in residential occupancies, therefore this study focused on fire patterns on painted gypsum wallboard which is a common interior finish for residential occupancies.  A common height for a residential occupancies is 2.4~m in height.  With a wall height of 2.4~m, these experiments used small fires with flame heights of approximately 1.2~m.  The resulting patterns on the gypsum board wall would be representative of a pre-flashover fire pattern.

@@

This study includes several series of real-scale, replicate fire experiments to examine the repeatability of pre-flashover fire patterns and, ultimately, examine the ability of simple correlations and the Fire Dynamic Simulator to recreate characteristics of the fire patterns.

@@

Statistical methods were used to guide the wall fire and corner fire experiments.  The analysis ensured that the appropriate number of replicate experiments were conducted to provide an assessment of burn pattern repeatability within a given level of uncertainty.  Since the coefficient of variation (COV) may be different for a given type of experiment, the number of replicates required would be larger for experiments with a larger COV in order obtain a valid assessment of repeatability.

@@

A free standing, steel frame with non-combustible, calcium silicate faced wall was instrumented to measure the thermal exposure of the wall from the fire in terms of temperature and heat flux.  The fires were initially placed at predetermined, fixed positions away from the wall. The burner was moved closer toward the wall until the burner was against the wall.

@@

A free standing, steel frame with a panel of gypsum wallboard attached to the surface was instrumented to measure the thermal exposure of the wall from the fire in terms of temperature and heat flux.  The fires were initially placed at predetermined, fixed positions away from the wall. The burner was moved closer toward the wall until the burner was against the wall.  If the results indicated that no burn pattern would result from a given combination of fuel and position, then that fuel and position combination were eliminated from the remainder of the test matrix.

@@

These experiments used the same test compartment as the wall fire pattern experiments.  In these cases, the fire was located against both walls in a corner of the compartment.  The experimental process was the same as the wall fire pattern experiments.

@@

Cone calorimeter experiments were conducted, based on ASTM~E1354, to estimate the ignition temperature, the time to ignition and the heat release rate from samples of painted gypsum wallboard under thermal flux conditions ranging from 25~kW/m$^2$ to 75~kW/m$^2$ ~\cite{ASTM_E1354}.

@@

FDS~\cite{McGrattan:2014} and Smokeview~\cite{Forney:2014} were used to examine the ability to numerically reproduce the source fires and the instrumented wall experiments. These simulation tools provide results in the form of visual field data.  In other words they provide results in the same format that the investigator would see or document the fire pattern.  Comparison between the experiments and the model may provide valuable validation data for the model.

@@

Point measurements were reduced and compared with values from the literature on fire plumes and their interaction with vertical surfaces.  The heat release rates, heat flux measurements, and flame and surface temperatures were used as input for heat transfer analysis to develop empirical coefficients for the specific experimental arrangements, which would be useful in the model recreations.
The key measurements for repeatability of the fire patterns on the gypsum wallboard are height, maximum width, shape, and area.  The two significant outcomes will be the repeatability of the fire patterns themselves and the ability to numerically recreate key measures of the pattern.

@@

A series of experiments were conducted to characterize a limited set of  residential scale, pre-flashover, or early growth stage, fires.  A fixed fire source footprint, a 30.5~cm square, was chosen for this study and a range of common fuels that could be found in and around a residence were selected.  The fires were positioned under an oxygen consumption calorimetry hood to determine the heat release rate.  Heat release rate, flame/plume gas temperatures, and total heat flux from the flames were measured and flame movement and height were recorded with photographs and videos.  Given the focus on residential fires, small source fires were selected so that the burn pattern would be limited to less than 2.4~m in height.  Replicate experiments were conducted with each of the fuels, in order to examine the repeatability of the fires and to quantify the variability in the flames.

@@

Three different fuels were used in this study.  Experiments were conducted with a natural gas fueled burner, a liquid fueled pan fire, and a solid fuel. Natural gas was chosen as a fuel for two reasons; it is a common fuel used for heating homes and it is used for calibrating the oxygen depletion calorimeters in NIST’s Large Fire Laboratory.  Therefore the heat content of the gas was monitored and mass flow controlled to provide a heat release rate based on fuel mass flow for comparison with the heat release rate values measured with the calorimeter.  The gas burner arrangement provided the most idealized fire, in that it could be ignited and brought up to a near steady state heat release rate within a few seconds and then be shut down just as quickly.  This provided an approximate `square wave' exposure to the gypsum wallboard.  This is important for determining the potential amount of energy transferred to the exposed surface.

The natural gas burner measured 30.5~cm on each side and the top surface of the burner was elevated 7.5~cm above the floor. The shell of the burner was made from steel and was filled with ``pea sized'' gravel with an average diameter of 6~mm.  A steady state heat release rate of approximately 80 kW was used for the experiments.

Gasoline was the fuel of choice for the liquid fueled pan fire.  Gasoline is not an ideal fuel to use in experiments because it is a mixture of components and additives that are specific to manufacturers. In addition, the mixtures change from season to season, and vary based on requirements or restrictions of the locality.  However, fire incident data shows that gasoline was the first item ignited in the majority of intentional fires where flammable or combustible liquids were used~\cite{Rohr:2001}. In addition, data from forensic laboratories, collected by Babrauskas and others, indicates that gasoline is the most prevalent accelerant found as part of the analysis of fire debris samples~\cite{Babrauskas:2003,Chasteen:2010}.

The gasoline pan fire used 500~mL of unleaded, 87 octane gasoline.  The pan was constructed from 6~mm thick steel.  The inner dimensions of the pan were 30.5~cm on each side and 1.9~cm high.  The pan was elevated in order to bring the pan lip height up to 7.5~cm above the floor to be the same height as the natural gas burner.  The 500~mL of gasoline had a fuel depth of 5.5~mm and provided a burn time of approximately 300~s.

Flexible polyurethane foam was chosen as the solid fuel. According to the Polyurethane Foam Association, flexible polyurethane foam is the most widely used cushioning material in upholstered furnishing and mattresses. More than 1.2 billion pounds are produced in the U.S. every year~\cite{Foam,Polyurethane_Foam}. In studies by Ahrens and Evarts of the NFPA, upholstered furniture, and mattresses and bedding were the first items ignited in 33~\% of the fatal fires in the U.S. based on averaged data from 2005 through 2009~\cite{Ahrens:2011,Evarts:2011}.

The flexible polyurethane foam had a density of approximately 23~kg/m$^3$.  Each block of foam was 30.5~cm on each side and was 7.5~cm thick. The bottom and sides of the foam block were wrapped in aluminum foil to prevent the fuel from flowing or spreading as it burned to ensure that the fuel area remained with the same fixed area as the gasoline and natural gas.

@@

On the averaged heat release rate curve for each fuel, the shaded area surrounding the average represents the 95~\% confidence level (2$\sigma$), based on Type A evaluation of the uncertainty is provided~\cite{Taylor:1994}.  A Type A evaluation of the uncertainty is any statistical analysis of the data from a series of observations.  In this case the evaluation assumed a normal distribution of the data such that doubling the calculated standard deviation (2$\sigma$) would yield a confidence level of approximately 95~\%.

@@

the data is plotted in the style used in FDS Validation

continuous flame region 
intermittant flame region
plume region

@@

steep thermal gardiant
